---
title: "ReproResearch_Project1"
author: "Adam Winter"
date: "March 23, 2018"
output: md_document
---
This small chunk is for saving away image files that are produced.   
```{r opts, echo = TRUE}
knitr::opts_chunk$set(
  fig.path = "figures/")
```
#Read in the activity data set
```{r}
setwd("/Users/adamwinter/R_Analytics/Data/")
data <- read.csv("activity.csv")
str(data)
head(data)
```
#Find mean total number of steps taken per day
First add up steps on each day  
```{r}
stepTotals <- aggregate(data$steps ~ data$date, FUN = sum)
colnames(stepTotals) <- c("date","steps")
```
Then, make a histogram with the totals  
```{r}
hist(stepTotals$steps, breaks = 5, col = 4, axes = TRUE, 
     main = "Total Number of Steps", xlab = "Number of Steps")
```
Calculate mean and median of the total number of steps taken per day  
```{r}
avg <- mean(stepTotals$steps)
med <- median(stepTotals$steps)
#values are displayed in the text below
```
The mean is `r avg` and the median is `r med`  

#What is the average daily activity pattern?  
Collapse data to a new data frame with the inteval and average number of steps  
```{r}
intervals <- aggregate(data$steps ~ data$interval,FUN = mean)
colnames(intervals) <- c("interval","steps")
```
Graph the time series plot
```{r}
plot(intervals$interval,intervals$steps, type = "l", col = 1,
     main = "Average number of steps across all days", 
     xlab = "Intervals", ylab = "Average number of steps")
```
Find interval with maximum average number of steps
```{r}
maxSteps <-sort(intervals$steps, decreasing = TRUE)[1]
maxInt <- as.numeric(intervals[intervals$steps == maxSteps,][1])
#The value is embeded and displayed in the text below
```
The 5-minute interval at `r maxInt` contains the maximum number of steps of `r maxSteps`.    

#Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r}
#sum(is.na(data$date)) and sum(is.na(data$interval)) show
#that the only column with NA values is data$steps
numNA <- sum(is.na(data$steps))
```
The number of rows with NAs is `r numNA`.

Now, I create a new dataset that is equal to the original dataset but with the missing data filled in.  I replace NA with the average value for the given 5-minute interval across all the days.  Average steps for each 5-minute interval are already caluclated in `intervals`.  
```{r}
fillData <- data
for(i in 1:nrow(data))
{
    if(is.na(data$steps[i]))    #row is missing steps
    {
        #get the interval of that row
        interval_val <- data$interval[i]
        #find the row in intervals with the avg steps for interval_val
        row_val <- which(intervals$interval == interval_val)
        #find the number of steps in that row
        avgSteps <- intervals$steps[row_val]
        #replace the NA value
        fillData$steps[i] <- avgSteps
    }
}
#Check that we got all of them by sum(is.na(fillData$steps)) == 0
```
Add up steps on each day.  
```{r}
fillStepTotals <- aggregate(fillData$steps ~ fillData$date, FUN = sum)
colnames(fillStepTotals) <- c("date","steps")
```
Make a histogram of the total number of steps taken each day.  
```{r}
hist(fillStepTotals$steps, breaks = 5, col = 4, axes = TRUE, 
    main = "Total Number of Steps (Filled In)", xlab = "Number of Steps")
```
Calculate mean and median steps for the new filled in data.  
```{r}
fillAvg <- mean(fillStepTotals$steps)
fillMed <- median(fillStepTotals$steps)
#values are displayed in the text below
```
The new filled-data mean is `r fillAvg` and the new filled-data median is `r fillMed`.  
The old incomplete data mean was `r avg` and the median was `r med`.  

Therefore, filling in the NA step values with the average number of steps in that 5-minute interval had the effect of making the median equal the mean.  The NA values were not spred throughout the data but rather isolated to every interval on a few select days.  When the values were filled, thier sum was equal to the mean and caused the curve to be more centralized, which caused a shift in the median.  

#Are there differences in activity patterns between weekdays and weekends?
The `weekdays()` function is helpful for this part.  It takes a date and returns the day of the week as a string.  We are using the filled dataset without missing values.  
```{R}
 #Create a vector with the day of the week for each date.
day <- weekdays(as.Date(fillData$date))
dayType <- vector()
for(i in 1:nrow(fillData))  #fill the dayType vector based on day
{
    if(day[i] == "Saturday" || day[i] == "Sunday")
        dayType[i] <- "weekend"
    else
        dayType[i] <- "weekday"
}

fillData$dayType <- dayType
# make new column a factor variable with levels of "weekday" and "weekend"
fillData$dayType <- factor(fillData$dayType)  
```
Create a table that has the average number of steps taken, averaged across all weekday days or weekend days, at each interval.  
```{r}
intervalsByDay <- aggregate(steps ~ interval + dayType, data = fillData, FUN = mean)
#Take a look at the new table
head(intervalsByDay)
```
Make a panel plot containing a time series plot of the average numeber of steps at each interval of that dayType.  I am using the lattice plotting system.
```{r}
library(lattice)
xyplot(steps ~ interval | dayType, data = intervalsByDay, type = "l", layout = c(1,2), xlab = "Interval", ylab = "Number of Steps")
```









